---
title: "Fitting the STAGE Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting the STAGE Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

library(stage)
library(tidyverse)
theme_set(theme_bw())
set.seed(1)
```

This vignette shows how to:

- fit the STAGE model to a simple simulated dataset,
- extract posterior transition points, and
- extend to a hierarchical (multi-population) model.

# Simulating data

To demonstrate the workflow, we simulate a simple dataset of lengths in \([L, U]\) with a smooth transition
around a true \(m_{50}\).

```{r simulate-data}
N <- 200
L <- 1000
U <- 1500
true_m50 <- 1250
true_d   <- 100

# Simulate covariate
x <- runif(N, L, U)

# Smooth transition in P(mature)
p <- plogis((x - true_m50) / (true_d / 4))

# Binary maturity indicator
y <- rbinom(N, 1, p)

head(tibble(x = x, y = y))
```


# Fitting the STAGE model

We now fit the STAGE model using `fit_stage()`.  
Because STAGE is a Bayesian model fit via Stan, we keep `chains` and `iter` small in this vignette to keep
runtime acceptable; for real applications you should increase these.

```{r fit-stage, eval=FALSE}
fit <- fit_stage(x, y, L = L, U = U, chains = 2, iter = 1000)
fit
```

You can inspect the posterior with standard tools such as `summary(fit$fit)`, `bayesplot`, or `posterior` utilities
if you want more diagnostics.


## Extracting the transition point

The helper `transition_point()` returns posterior summaries of the global transition point \(m_{50}\).

```{r transition-point, eval=FALSE}
transition_point(fit)
```

Typical output is a small data frame with posterior mean, median, and credible intervals for `m50`.


## Predicting maturity probabilities

We can also predict \(\Pr(\text{mature} \mid x)\) over a grid of lengths, and plot the posterior mean transition curve.

```{r prediction, eval=FALSE}
x_grid <- seq(L, U, length.out = 200)

p_hat <- predict(fit, x_grid, type = "prob")

plot(
  x_grid, p_hat, type = "l",
  xlab = "Length (x)",
  ylab = "P(mature)",
  las = 1
)
```


# Hierarchical (multi-population) model

In many applications we have multiple populations (e.g. regions, stocks, or sexes) and we expect:

- a **shared average transition point** \(\mu_{m50}\), and
- **population-specific deviations**.

STAGE implements this using a hierarchical structure
\[
m50_{\text{pop}[j]} = \mu_{m50} + z_j \,\sigma_\alpha,
\]
with a non-centred parameterisation for efficient sampling.

A simple two-group example:

```{r hierarchical-example, eval=FALSE}
group <- rep(c("A", "B"), each = N/2)

fit2 <- fit_stage(
  x, y,
  group = group,
  L = L, U = U,
  chains = 2, iter = 1000
)

transition_point(fit2)
```

`transition_point(fit2)` returns both:

- a global `m50` (overall mean transition)  
- group-specific `m50_pop[j]` values.


# Summary

In this vignette we:

- introduced the plateau–Gaussian likelihood underlying STAGE,
- visualised the class densities and resulting transition probabilities,
- simulated a simple dataset,
- fit the STAGE model to estimate the transition point, and
- extended the model to a hierarchical multi-population setting.

In practice, you will want to:

- choose \([L, U]\) to capture the transition zone plus some plateau on each side,
- check posterior diagnostics (divergences, R-hat, effective sample sizes),
- and interpret \(m_{50}\) and its uncertainty in the biological context.

Future vignettes will expand on:

- comparing STAGE to logistic regression and Gaussian LDA, and
- sensitivity to the choice of \([L, U]\) and prior settings.
