---
title: "Fitting the STAGE Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting the STAGE Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

library(stage)
library(tidyverse)
theme_set(theme_bw())
set.seed(1)
```

# Fitting the STAGE model

This vignette shows how to:

- visualise the STAGE likelihood,
- fit the STAGE model to a simple simulated dataset,
- extract posterior transition points, and
- extend to a hierarchical (multi-population) model.

Conceptually, STAGE is a Bayesian generative classifier for estimating a transition point (e.g. length at maturity).  

Each class is fit with an asymmetric plateau–Gaussian density:

- the immature class has a **uniform plateau** followed by a **Gaussian tail down into the transition**,  
- the mature class has a **Gaussian tail up out of the transition** followed by a **uniform plateau**.

This construction focuses inference on the overlap region and prevents distant observations 
(e.g. very small immatures, very large matures) from unduly influencing the transition point. 


## The STAGE likelihood

For a given truncation interval \([L, U]\) and transition point \(m_{50}\),

- the immature class (\(y = 0\)) has mean \(\mu_0 = m_{50} - d/2\),
- the mature class (\(y = 1\)) has mean \(\mu_1 = m_{50} + d/2\),

with a shared Gaussian spread \(\sigma_x\) in the transition region and a plateau outside:

- **Immature (left) density**  
  uniform over \([L, \mu_0]\), Gaussian for \(x > \mu_0\)
- **Mature (right) density**  
  Gaussian for \(x < \mu_1\), uniform over \([\mu_1, U]\)

The **class-conditional densities** are normalised so each integrates to 1, and the transition probability
\(\Pr(y=1 \mid x)\) arises from **Bayes’ rule** by comparing the two densities.


### Visualising a single STAGE transition

We can now visualise a simple STAGE configuration, showing:

- the immature and mature densities, and
- the resulting transition probability \(\Pr(y=1 \mid x)\).


```{r visualize-likelihood, fig.width=6, fig.height=4}
# parameters for the illustration
L      <- 1000
U      <- 1700
sigma  <- 120
m50    <- 1350           # mid point
d      <- 300            # separation
mu0    <- m50 - d/2      # 1200
mu1    <- m50 + d/2      # 

pdat <- tibble(x = L:U) |> 
  rowwise() |> 
  mutate(
    "y=0" = plateau_gaussian_density_left(
      x, mu = mu0, sigma = sigma, L = L),
    "y=1" = plateau_gaussian_density_right(
      x, mu = mu1, sigma = sigma, U = U)
  ) |> 
  pivot_longer(
    cols = c(`y=0`, `y=1`),
    names_to = "Class",
    values_to = "Density"
  )


# base plot
ggplot(pdat) + 
  aes(x=x, y=Density, fill=Class) +
  geom_area(position = "identity", alpha = 0.5)

# maximum density
maxd <- max(pdat$Density)

p <- ggplot(pdat) +
  aes(x = x, y = Density, fill = Class) +
  geom_area(alpha = 0.45, position = "identity") +
  scale_fill_manual(values = c("#00BFC4", "#F8766D")) +
  labs(
    x = "x",
    y = "Density",
    fill = "State"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# add labelled structure
p +
  # vertical lines for key parameters
  geom_segment(aes(x = mu0, xend=mu0, y=0, yend=maxd), linetype = 2, colour = "#00BFC4") +
  geom_segment(aes(x = mu1, xend=mu1, y=0, yend=maxd), linetype = 2, colour = "#F8766D") +
  geom_segment(aes(x = m50, xend=m50, y=0, yend=maxd),  linetype = 2, colour = "black") +
  geom_vline(xintercept = L,    linetype = 3, alpha = 0.3) +
  geom_vline(xintercept = U,    linetype = 3, alpha = 0.3) +

  # annotate regions
  annotate("text", x = L + 20,  y = 0.003, label = "Plateau (y = 0)", colour = "#00BFC4", hjust = 0) +
  annotate("text", x = U - 20,  y = 0.003, label = "Plateau (y = 1)", colour = "#F8766D", hjust = 1) +

  annotate("text", x = mu0, y = maxd, label = expression(mu[0]), colour = "#00BFC4", vjust = -0.4) +
  annotate("text", x = mu1, y = maxd, label = expression(mu[1]), colour = "#F8766D", vjust = -0.4) +
  annotate("text", x = m50, y = maxd, label = expression(m[50]), vjust = -0.5) +

  # illustrate d
  annotate("segment",
           x = mu0, xend = mu1, y = 0.001, yend = 0.001,
           arrow = arrow(ends = "both", length = unit(0.12, "cm"))) +
  annotate("text", x = m50*1.03, y = 0.001*1.05, label = "d", vjust = 0, hjust = 0)
```



# Simulating data

To demonstrate the workflow, we simulate a simple dataset of lengths in \([L, U]\) with a smooth transition
around a true \(m_{50}\).

```{r simulate-data}
N <- 200
L <- 1000
U <- 1500
true_m50 <- 1250
true_d   <- 100

# Simulate covariate
x <- runif(N, L, U)

# Smooth transition in P(mature)
p <- plogis((x - true_m50) / (true_d / 4))

# Binary maturity indicator
y <- rbinom(N, 1, p)

head(tibble(x = x, y = y))
```


# Fitting the STAGE model

We now fit the STAGE model using `fit_stage()`.  
Because STAGE is a Bayesian model fit via Stan, we keep `chains` and `iter` small in this vignette to keep
runtime acceptable; for real applications you should increase these.

```{r fit-stage, eval=FALSE}
fit <- fit_stage(x, y, L = L, U = U, chains = 2, iter = 1000)
fit
```

You can inspect the posterior with standard tools such as `summary(fit$fit)`, `bayesplot`, or `posterior` utilities
if you want more diagnostics.


## Extracting the transition point

The helper `transition_point()` returns posterior summaries of the global transition point \(m_{50}\).

```{r transition-point, eval=FALSE}
transition_point(fit)
```

Typical output is a small data frame with posterior mean, median, and credible intervals for `m50`.


## Predicting maturity probabilities

We can also predict \(\Pr(\text{mature} \mid x)\) over a grid of lengths, and plot the posterior mean transition curve.

```{r prediction, eval=FALSE}
x_grid <- seq(L, U, length.out = 200)

p_hat <- predict(fit, x_grid, type = "prob")

plot(
  x_grid, p_hat, type = "l",
  xlab = "Length (x)",
  ylab = "P(mature)",
  las = 1
)
```


# Hierarchical (multi-population) model

In many applications we have multiple populations (e.g. regions, stocks, or sexes) and we expect:

- a **shared average transition point** \(\mu_{m50}\), and
- **population-specific deviations**.

STAGE implements this using a hierarchical structure
\[
m50_{\text{pop}[j]} = \mu_{m50} + z_j \,\sigma_\alpha,
\]
with a non-centred parameterisation for efficient sampling.

A simple two-group example:

```{r hierarchical-example, eval=FALSE}
group <- rep(c("A", "B"), each = N/2)

fit2 <- fit_stage(
  x, y,
  group = group,
  L = L, U = U,
  chains = 2, iter = 1000
)

transition_point(fit2)
```

`transition_point(fit2)` returns both:

- a global `m50` (overall mean transition)  
- group-specific `m50_pop[j]` values.


# Summary

In this vignette we:

- introduced the plateau–Gaussian likelihood underlying STAGE,
- visualised the class densities and resulting transition probabilities,
- simulated a simple dataset,
- fit the STAGE model to estimate the transition point, and
- extended the model to a hierarchical multi-population setting.

In practice, you will want to:

- choose \([L, U]\) to capture the transition zone plus some plateau on each side,
- check posterior diagnostics (divergences, R-hat, effective sample sizes),
- and interpret \(m_{50}\) and its uncertainty in the biological context.

Future vignettes will expand on:

- comparing STAGE to logistic regression and Gaussian LDA, and
- sensitivity to the choice of \([L, U]\) and prior settings.
